Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-03-09 manuscript/climate_sensitivity_SI.Rmd
misc cleanup to conform to journal guidelines
#82

---
title: ""
author: ""
output:
  pdf_document:
    fig_caption: yes
biblio-style: apalike
csl: apa.csl
header-includes:
- \usepackage{booktabs}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
---

>>>>

---
title: ""
author: ""
output:
  pdf_document:
    fig_caption: yes
biblio-style: apalike
csl: apa.csl
header-includes:
- \usepackage{booktabs}
- \usepackage{caption}
- \usepackage{setspace}\onehalfspacing
- \captionsetup[figure]{labelformat=empty}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
---

### 2020-03-09 manuscript/climate_sensitivity_draft.Rmd
misc cleanup to conform to journal guidelines
#82

---
title: ""
author: ""
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: ["book.bib",  "packages.bib"]
biblio-style: "apalike"
csl: apa.csl
header-includes:
- \usepackage{setspace}\onehalfspacing
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Figure S1). Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies were published in association with @helcoski_growing_2019 (DOI: 10.5281/ zenodo.2649302).

@@

Crown position--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999. While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of our hypothesis (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative. An analysis of crown position relative to height (Fig. 2d) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among canopy positions (Fig. S3), with average tree height growth confined to ~0.82 m from 1966 to 1977, ~1.45 m from 1977 to 1999, and ~1.97 m from 1999 to 2018. However, dominant and co-dominant trees were similar in height (Figs. 2d, S3).

Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Figure S1) [@R-dynatopmodel]. Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006]. TWI calculation depends on an input of a digital elevation model (DEM; ~3.7 m resolution from the elevatr package [@R-elevatr]), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely. From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from the stream.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth, *i.e.*, droughts that substantially impacted the forest community. We identified three drought years: 1966, 1977, and 1999 (Figs. 1, S2, Table S3). These were the years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site. PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. These were also years with widespread tree growth reduction ("pointer years"), here defined as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011]. $Rt$ values <1 and >1 indicate growth reductions and increases, respectively. Pointer years were identified using the pointRes package [@R-pointRes] in R. In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3). Rather, the severity of growth reduction could probably be explained in large part by defoliation by gypsy moths (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp. in the area from approximately 1988 through 1995 [@twery_effects_1991].

The droughts differed in intensity and antecedent moisture conditions (Fig. S2, Table S3). 
The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts. 
The 1977 drought was the least intense throughout the growing season, and it was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July (-4.53).

@@

At the community-level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig. 1b). In each drought, roughly 30% of the cored trees had $Rt \le 0.7$: 29% in 1966, 32% in 1977, and 27% in 1999. However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.

@@

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. 2d). When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$. Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).  When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed trees the highest. In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. 2). Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. 2a). Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. 2b). Air temperature did not vary across the vertical profile (Fig. 2c).

@@

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. 1a, @helcoski_growing_2019) and had substantial impacts on tree growth (Fig. 1b). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. S2), severely stressed a larger portion of trees (Fig. 1b). The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. 1b), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).

Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, the collinearity between the two variables (Fig. 2d) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. 2d) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. 2a-b). Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig. 2d) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5). Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the vulnerability conferred by height--and, in fact, greater moisture access in non-droought years (here, higher TWI) appears to make trees more vulnerable to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

>>>>

---
title: ""
author: ""
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: ["book.bib",  "packages.bib"]
biblio-style: "apalike"
csl: apa.csl
header-includes:
- \usepackage{setspace}\onehalfspacing
- \usepackage{lineno}
- \linenumbers
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

We analyzed tree-ring data (cambial growth increment) from 571 trees representing the twelve species with the greatest contributions to woody aboveground net primary productivity ($ANPP_{stem}$), which together comprised 97% of study plot $ANPP_{stem}$ between 2008 and 2013 [@helcoski_growing_2019] (Fig. **S1**). Cores were collected within the ForestGEO plot at breast height (1.3m) in 2010-2011 or 2016-2017. In 2010-2011, cores were collected from randomly selected live trees of each species that had at least 30 individuals $\ge$ 10 cm DBH [@bourg_initial_2013]. In 2016-2017, cores were collected from all trees found dead during annual mortality censuses [@gonzalezakre_patterns_2016]. Cores were sanded, measured, and crossdated using standard procedures, as detailed in [@helcoski_growing_2019]. The resulting chronologies were published in association with @helcoski_growing_2019 (DOI: 10.5281/ zenodo.2649302).

@@

Crown position--a categorical variable including dominant, co-dominant, intermediate, and suppressed--was recorded for all cored trees that remained standing during the growing season of 2018 following the protocol of @jennings_assessing_1999. While some tree crowns undoubtedly changed position over the past several decades, in this case the bias would be unlikely to result in false acceptance of our hypothesis (*i.e.*, type I error unlikely, type II error possible), making our hypothesis test conservative. An analysis of crown position relative to height (Fig. **2d**) and height changes since the beginning of the study period indicated that changes between focal drought years (1966, 1977, and 1999; see below) were fairly small relative to differences among canopy positions (Fig. **S3**), with average tree height growth confined to ~0.82 m from 1966 to 1977, ~1.45 m from 1977 to 1999, and ~1.97 m from 1999 to 2018. However, dominant and co-dominant trees were similar in height (Figs. **2d**, **S3**).

Topographic wetness index (TWI) was calculated using the dynatopmodel package in R (Fig. **S1**) [@R-dynatopmodel]. Originally developed by @beven_physically_1979, TWI was part of a hydrological run-off model and has since been used for a number of purposes in hydrology and ecology [@sorensen_calculation_2006]. TWI calculation depends on an input of a digital elevation model (DEM; ~3.7 m resolution from the elevatr package [@R-elevatr]), and from this yields a quantitative assessment defined by how “wet” an area is, based on areas where run-off is more likely. From our observations in the plot, TWI performed better at categorizing wet areas than the Euclidean distance from the stream.

@@

We identified droughts within the time period 1950-2009, defining drought [@slette_how_2019] as events with both anomalously dry peak growing season climatic conditions and widespread reductions in tree growth, *i.e.*, droughts that substantially impacted the forest community. We identified three drought years: 1966, 1977, and 1999 (Figs. **1**, **S2**, Table S3). These were the years with the lowest Palmer Drought Severity Index (PDSI) during May-August (MJJA; Table S3), which were identified by @helcoski_growing_2019 as the months of the current year to which annual tree growth was most sensitive at this site. PDSI divisional data for Northern Virginia were obtained from NOAA (https://www7.ncdc.noaa.gov/CDO/CDODivisionalSelect.jsp) in December 2017. These were also years with widespread tree growth reduction ("pointer years"), here defined as those where >25% of the cored trees experienced >30% reduction in basal area increment ($BAI$) relative to the previous 5 years, following the drought resistance ($Rt$) metric of [@lloret_components_2011]. $Rt$ values <1 and >1 indicate growth reductions and increases, respectively. Pointer years were identified using the pointRes package [@R-pointRes] in R. In addition to the focal drought years, 1991 also met this critera (26.5% of trees experienced >30% growth reduction, mean resistance= -13.8%) but was excluded because it was not among the driest of the time period (Table S3). Rather, the severity of growth reduction could probably be explained in large part by defoliation by gypsy moths (*Lymantria dispar* L.), which was documented to have strongly impacted *Quercus* spp. in the area from approximately 1988 through 1995 [@twery_effects_1991].

The droughts differed in intensity and antecedent moisture conditions (Fig. **S2**, Table S3). 
The 1966 drought was preceded by two years of moderate drought during the growing season and severe to extreme drought starting the previous fall and in August reached the lowest growing season $PDSI$ (-4.82) of the three droughts. 
The 1977 drought was the least intense throughout the growing season, and it was preceded by 2.5 years of near-normal conditions, making it the mildest of the three droughts. 
The 1999 drought was preceded by wetter than average conditions until the previous June, but reached the lowest PDSI during May-July (-4.53).

@@

At the community-level, cored trees showed substantial growth reductions in all three droughts, with a mean $Rt$ of 0.86 in 1966 and 1999, and 0.84 in 1977  (Fig. **1b**). In each drought, roughly 30% of the cored trees had $Rt \le 0.7$: 29% in 1966, 32% in 1977, and 27% in 1999. However, some individuals exhibited increased growth, *i.e.*, $Rt > 1.0$: 26% of trees in 1966, 22% in 1977, and 26% in 1999.

@@

Crown position varied as expected with height (dominant > co-dominant > intermediate > suppressed), but with substantial variation (Fig. **2d**). When considered alone, crown position had a significant response only in the 1966 drought, during which trees with dominant crown position had the lowest $Rt$. Crown position was a much poorer predictor of $Rt$ than was height in the single-variable tests (Table 4), lending little overall support to the hypothesis that crown exposure reduces $Rt$ (Table 1).  When height was included in the model, crown position was a significant predictor in the 1999 drought, with lowest $Rt$ for suppressed and then intermediate trees. Crown position was included in some of the full models (Table 5). In 1977, where height was not included in the full model, dominant trees had the lowest $Rt$, and suppressed trees the highest. In contrast, in full models including both height and crown position (all droughts and 1999), the lowest $Rt$ was in suppressed, followed by intermediate, trees.

In the years for which we have vertical profiles in climate data (2016-2018), taller trees--or those in dominant crown positions-- were generally exposed to higher evaporative demand during the peak growing season months (May-August; Fig. **2**). Specifically, maximum daily wind speeds were significantly higher above the top of the canopy (40-50m) than within and below (10-30m) (Fig. **2a**). Relative humidity was also somewhat lower during June-August, ranging from ~50-80% above the canopy and ~60-90% in the understory (Fig. **2b**). Air temperature did not vary across the vertical profile (Fig. **2c**).

@@

The droughts considered here were of a magnitude that has occurred with an average frequency of approximately once every 10-15 years (Fig. **1a**, @helcoski_growing_2019) and had substantial impacts on tree growth (Fig. **1b**). These droughts were classified as severe (1977) or extreme (1966, 1999) according to the PDSI metric and have been linked to tree mortality in the eastern United States [@druckenbrod_redefining_2019]; however, extreme, multiannual droughts or so-called “megadroughts” of the type that have triggered massive tree die-off in other regions (e.g., @allen_global_2010; @stovall_tree_2019) have not occurred in the Eastern United States within the past several decades [@clark_impacts_2016]. Of the droughts considered here, the 1966 drought, which was preceded by two years of dry conditions (Fig. **S2**), severely stressed a larger portion of trees (Fig. **1b**). The tendency for large trees to have lowest resistance was most pronounced in this drought, consistent with other findings that this physiological response increases with drought severity [@bennett_larger_2015; @stovall_tree_2019]. Across all three droughts, the majority of trees experienced reduced growth, but a substantial portion had increased growth (Fig. **1b**), potentially due to decreased leaf area of competitors during the drought. It is likely because of the moderate impact of these droughts, along with other factors influencing tree growth, that our best models characterize only a modest amount of variation: 11-13% for all droughts combined, and 21-26% for each individual drought (Table 5).

Our analysis indicates that tree height has a stronger influence on drought response than does canopy position (Tables 1,4,5). This is consistent with, and reinforces, previous findings that biophysical constraints make it impossible for trees to efficiently transport water to great heights and simultaneously maintain strong resistance and resilience to drought-induced embolism [@olson_plant_2018; @couvreur_water_2018; @roskilly_conflicting_2019]. However, the collinearity between the two variables (Fig. **2d**) makes it impossible to confidently partition causality. Taller trees are more likely to be in dominant canopy positions (Fig. **2d**) and, largely as a consequence of their position relative to others, face different microenvironments (Fig. **2a-b**). Even under non-drought conditions, evaporative demand and maximum leaf temperatures increase with tree height [@smith_temperature_1977; @bretfeld_plant_2018; @kunert_revised_2017], and such conditions would incur additional stress during drought, when solar radiation tends to be higher and less water is available for evaporative cooling of the leaves. However, some decoupling between height and canopy position is introduced by the configuration of neighboring trees (Fig. **2d**) [@muller-landau_testing_2006], and height was an overall stronger predictor of drought response than crown position (Tables 1,4,5). Belowground, taller trees would tend to have larger root systems, but the potentially greater access to water did not override the vulnerability conferred by height--and, in fact, greater moisture access in non-droought years (here, higher TWI) appears to make trees more vulnerable to drought [@zuleta_drought-induced_2017; @stovall_tree_2019].

### 2020-03-09 manuscript/tables&figures.Rmd
added climate ref
misc cleanup to conform to journal guidelines
#82

---
title: ""
author: ""
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: ["book.bib",  "packages.bib"]
biblio-style: "apalike"
csl: apa.csl
header-includes:
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

@@

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species. Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b) show the distribution of resistance values for each drought.](tables_figures/Figure1.png){width=500px}

>>>>

---
title: ""
author: ""
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: ["book.bib",  "packages.bib"]
biblio-style: "apalike"
csl: apa.csl
header-includes:
- \usepackage{float}
- \usepackage{booktabs}
- \usepackage{pdflscape}
- \usepackage{setspace}\onehalfspacing
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage{caption}
- \captionsetup[table]{font=small}
- \captionsetup[figure]{font=small}
- \captionsetup[table]{labelformat=empty}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{dcolumn} #for sidewaystable
---

@@

\newpage
### Figure Legends

**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (**a**) shows peak growing season (May-August) climate conditions and residual chronologies for each species. PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (**b**) show the distribution of resistance values for each drought.

**Figure 2. Height profiles in growing season climatic conditions, tree heights by crown position, and leaf hydraulic traits** The top row shows averages ($\pm$ SD) of daily maxima and minima of (**a**) wind speed, (**b**) relative humidity ($RH$), and (**c**) air temperature ($T_{air}$) averaged over each month of the peak growing season (May-August) from 2016-2018. In these plots, heights are slightly offset for visualization purposes. Also shown is (**d**) 2018 tree heights by canopy position (see Table 2 for codes). In all plots, the dashed horizontal line indicates the 95th percentile of tree heigts in the ForestGEO plot.

@@

![**Figure 1. Climate and species-level growth responses over our study period, highlighting the three focal drougths (a) and community-wide responses** Time series plot (a) shows peak growing season (May-August) climate conditions and residual chronologies for each species. PET and PRE data were obtained from the Climatic Research Unit high-resolution gridded dataset (CRU TS v.4.01; Harris et al. 2014). Focal droughts are indicated by dashed lines, and shading indicates the pre-drought period used in calculations of the resistance metric. Figure modified from Helcoski *et al.* (2019). Density plots (b) show the distribution of resistance values for each drought.](tables_figures/Figure1.png){width=500px}
